---
title: "Rotation2"
author: "Wanru Lin"
date: "2023-01-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Read the data


```{r}
suppressPackageStartupMessages({
    library(dplyr)
    library(Seurat)
    library(patchwork)
    library(DropletUtils)
    library(ggplot2)
    library(pheatmap)
    library(reticulate)
    library(rjson)
    library(fishpond)
    library(HGNChelper)
    library(roe)
    library(SC3)
    library(scater)
})
```


```{r}
# return is cell for each row in the input count matrix
filter_empty <- function(unfiltered_counts) {
    br.out <- barcodeRanks(unfiltered_counts)
    e.out <- emptyDrops(unfiltered_counts)
    is.cell <- e.out$FDR <= 0.01
    is.cell[is.na(is.cell)] <- FALSE
    is.cell
}

create_seurat_object <- function(c, sce, min.cells = 0, min.feature = 0) {
    rownames(c) <- rownames(sce)
    colnames(c) <- colnames(sce)
    CreateSeuratObject(c, min.cells = min.cells, min.feature = min.feature)
}
```


```{r}
gid2name_df = read.csv("human2020A_geneid_to_name.tsv", header = FALSE, sep = "\t")
gid2name = gid2name_df$V2
names(gid2name) = gid2name_df$V1
```

```{r}
# pbmc_quant_dir = "/mnt/scratch3/alevin_fry_submission/Benchmark-snRNAseq/af-private-scripts/nf_pipeline/output/alevin_fry/pbmc10k/fry_unfilt_quant_usa_cr-like"
pbmc_sce_unfilt = FDL(1,output_format="raw")@sce
# get rid of the mitochondrial gene
pbmc_sce_unfilt = pbmc_sce_unfilt[rownames(pbmc_sce_unfilt) %in% names(gid2name), ]
# get overall count matrix
c <- pbmc_sce_unfilt@assays@data$spliced + pbmc_sce_unfilt@assays@data$unspliced + pbmc_sce_unfilt@assays@data$ambiguous

# filter empty drops
is.cell <- filter_empty(c)
pbmc_sce_filt = pbmc_sce_unfilt[,is.cell]
rownames(pbmc_sce_filt) = gid2name[rownames(pbmc_sce_filt)]
pbmc_sce_filt

```

```{r}
c_usa = pbmc_sce_filt@assays@data$spliced + pbmc_sce_filt@assays@data$ambiguous + pbmc_sce_filt@assays@data$unspliced
pbmc_usa <- create_seurat_object(c_usa, pbmc_sce_filt)
pbmc_usa
```


```{r}
pbmc_usa[["percent.mt"]] <- PercentageFeatureSet(pbmc_usa, pattern = "^MT-")
```

```{r}
VlnPlot(pbmc_usa, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```

```{r}
pbmc_usa <- subset(pbmc_usa, subset = nFeature_RNA > 500 & 
                   nFeature_RNA < 7000 & 
                   percent.mt < 25 & 
                   nCount_RNA < 40000)
pbmc_usa
```

```{r}
pbmc_sce_filt <- pbmc_sce_filt[, colnames(pbmc_usa)]
# pbmc_sce_filt <- pbmc_sce_filt[(rowSums(pbmc_sce_filt@assays@data$spliced + pbmc_sce_filt@assays@data$ambiguous)) != 0 & rowSums(pbmc_sce_filt@assays@data$unspliced) != 0, ]
```

```{r}
c_s = as.matrix(pbmc_sce_filt@assays@data$spliced)
colnames(c_s) = colnames(pbmc_sce_filt)
rownames(c_s) = rownames(pbmc_sce_filt)
```


```{r}
c_s_nozero1 = c_s[apply(c_s, 1, function(x) !all(x==0)),]
dim(c_s_nozero1)

c_s_nozero_t = t(c_s_nozero1)
dim(c_s_nozero_t)
```

## Highly variable gene selection


```{r}
# highly variable genes
hvg_c_s = VariableFeatures(seurat_object_s)
```

```{r}
c_s_hvg_counts = t(c_s[hvg_c_s,])
dim(c_s_hvg_counts)
```

## Calculate the nearest neighbor graph by manhattan or euclidean distance method for original data or data after highly variable gene selection

```{r}
library(KODAMA)
library(cccd)
knn_c_s_lp = list()
nng_c_s_lp = list()

knn_c_s_lp[[1]] = knn.dist(c_s_nozero_t, dist.meth = "manhattan",p=1)
nng_c_s_lp[[1]] = nng(dx = (knn_c_s_lp[[1]]),k = 30,method = "manhattan")

knn_c_s_lp[[2]] = knn.dist(c_s_nozero_t, dist.meth = "euclidean",p=2)
nng_c_s_lp[[2]] = nng(dx = (knn_c_s_lp[[2]]),k = 30,method = "euclidean")
```

```{r}
adj_mtx_manhattan = as_adjacency_matrix(nng_c_s_lp[[1]])
adj_mtx_euclidean = as_adjacency_matrix(nng_c_s_lp[[2]])

#knn_c_s_hvg_lp[[1]][1,which(adj_mtx_manhattan[1,]==0)] = 0
for(i in 1:689){
  knn_c_s_lp[[1]][i,which(adj_mtx_manhattan[i,]==0)] = 0
}
for(i in 1:689){
  knn_c_s_lp[[2]][i,which(adj_mtx_euclidean[i,]==0)] = 0
}

```




```{r}
knn_c_s_hvg_lp = list()
nng_hvg_lp = list()

knn_c_s_hvg_lp[[1]] = knn.dist(c_s_hvg_counts, dist.meth = "manhattan",p=1)
nng_hvg_lp[[1]] = nng(dx = (knn_c_s_hvg_lp[[1]]),k = 30,method = "manhattan")

knn_c_s_hvg_lp[[2]] = knn.dist(c_s_hvg_counts, dist.meth = "euclidean",p=2)
nng_hvg_lp[[2]] = nng(dx = (knn_c_s_hvg_lp[[2]]),k = 30,method = "euclidean")

```



```{r}
adj_mtx_manhattan = as_adjacency_matrix(nng_hvg_lp[[1]])
adj_mtx_euclidean = as_adjacency_matrix(nng_hvg_lp[[2]])



#knn_c_s_hvg_lp[[1]][1,which(adj_mtx_manhattan[1,]==0)] = 0
for(i in 1:689){
  knn_c_s_hvg_lp[[1]][i,which(adj_mtx_manhattan[i,]==0)] = 0
}
for(i in 1:689){
  knn_c_s_hvg_lp[[2]][i,which(adj_mtx_euclidean[i,]==0)] = 0
}

```

## Calculate the geodesic distance

```{r}
knn_spearman_cor_c_s = knn_spearman_cor(knn_c_s_lp[[1]],knn_c_s_lp[[2]])
knn_spearman_cor_c_s$spearman_cor

geod_manhattan_c_s = knn_spearman_cor_c_s$dist_knn_1
geod_euclidean_c_s = knn_spearman_cor_c_s$dist_knn_2
```

```{r}
knn_spearman_cor_2 = knn_spearman_cor(knn_c_s_hvg_lp[[1]],knn_c_s_hvg_lp[[2]])
knn_spearman_cor_2$spearman_cor

geod_manhattan_hvg = knn_spearman_cor_2$dist_knn_1
geod_euclidean_hvg = knn_spearman_cor_2$dist_knn_2
```

## Difference of geodesic distance

```{r}
diff_hvg = (geod_manhattan_hvg-geod_euclidean_hvg)
diff_c_s = (geod_manhattan_c_s-geod_euclidean_c_s)
diff_pca50 = (geod_manhattan_pca50 - geod_euclidean_pca50)
diff_pca30 = (geod_manhattan_pca-geod_euclidean_pca)

diff_m_hvg_vs_cs = (geod_manhattan_hvg-geod_manhattan_c_s)
diff_e_hvg_vs_cs = (geod_euclidean_hvg-geod_euclidean_c_s)

diff_m_pca30_vs_cs = geod_manhattan_pca-geod_manhattan_c_s
diff_e_pca30_vs_cs = geod_euclidean_pca-geod_euclidean_c_s
```


```{r}
df_diff_hvg = data.frame(diff_hvg_value = as.vector(diff_hvg))
df_diff_c_s = data.frame(diff_c_s_value = as.vector(diff_c_s))
df_diff_m_hvg_vs_cs = data.frame(diff_m_hvg_vs_cs_value = as.vector(diff_m_hvg_vs_cs))
df_diff_e_hvg_vs_cs = data.frame(diff_e_hvg_vs_cs_value = as.vector(diff_e_hvg_vs_cs))

df_diff_pca30 = data.frame(diff_pca30_value = as.vector(diff_pca30))

df_diff_m_pca30_vs_cs = data.frame(diff_m_pca30_vs_cs_value = as.vector(diff_m_pca30_vs_cs))
df_diff_e_pca30_vs_cs = data.frame(diff_e_pca30_vs_cs_value = as.vector(diff_e_pca30_vs_cs))
```


```{r}
df_diff_hvg_no = subset(df_diff_hvg,diff_hvg_value != 0)
df_diff_c_s = subset(df_diff_c_s,diff_c_s_value != 0)
df_diff_m_hvg_vs_cs = subset(df_diff_m_hvg_vs_cs,diff_m_hvg_vs_cs_value != 0)
df_diff_e_hvg_vs_cs = subset(df_diff_e_hvg_vs_cs,diff_e_hvg_vs_cs_value != 0)

df_diff_pca30 = subset(df_diff_pca30,diff_pca30_value != 0)

df_diff_m_pca30_vs_cs = subset(df_diff_m_pca30_vs_cs,diff_m_pca30_vs_cs_value != 0)
df_diff_e_pca30_vs_cs = subset(df_diff_e_pca30_vs_cs,diff_e_pca30_vs_cs_value != 0)
```

```{r}
median(df_diff_hvg$diff_hvg_value)
median(df_diff_c_s$diff_c_s_value)
median(df_diff_m_hvg_vs_cs$diff_m_hvg_vs_cs_value)
median(df_diff_e_hvg_vs_cs$diff_e_hvg_vs_cs_value)

median(df_diff_pca30$diff_pca30_value)
median(df_diff_m_pca30_vs_cs$diff_m_pca30_vs_cs_value)
median(df_diff_e_pca30_vs_cs$diff_e_pca30_vs_cs_value)
```



```{r}
library(ggplot2)

ggplot(data=df_diff_hvg,aes(diff_hvg_value))+ # (0,11373.91)
  geom_histogram(bins =100,fill="blue",color="black",alpha=0.3)+
  #geom_vline(aes(xintercept = mean(diff_hvg_value)))+
  geom_vline(aes(xintercept = median(diff_hvg_value)))+
  xlab("Difference of geodesic distance between using different ditance metric after filtering") 

ggplot(data=df_diff_c_s,aes(diff_c_s_value))+ # (0.00,27133.52)
  geom_histogram(bins =100,fill="blue",color="black",alpha=0.3)+
  geom_vline(aes(xintercept = median(diff_c_s_value)))+
  xlab("Difference of geodesic distance between using different ditance metric without filtering") 

ggplot(data=df_diff_m_hvg_vs_cs,aes(diff_m_hvg_vs_cs_value))+
  geom_histogram(bins =100,fill="blue",color="black",alpha=0.3)+
  geom_vline(aes(xintercept = median(diff_m_hvg_vs_cs_value)))+
  xlab("Difference of geodesic distance before and after filtering \n from KNN graph using Manhattan distance method") # (-23945,0)

ggplot(data=df_diff_e_hvg_vs_cs,aes(diff_e_hvg_vs_cs_value))+
  geom_histogram(bins =100,fill="blue",color="black",alpha=0.3)+
  geom_vline(aes(xintercept = median(diff_e_hvg_vs_cs_value)))+
  xlab("Difference of geodesic distance before and after filtering \n from KNN graph using Euclidean distance method") # (-1084.599,1110.909)

```



```{r}
# compare distance metrics in raw counts/ after filtering hvg/ after pca(30)
knn_spearman_cor(knn_c_s_hvg_lp[[1]],knn_c_s_hvg_lp[[2]])$spearman_cor
knn_spearman_cor(knn_c_s_lp[[1]],knn_c_s_lp[[2]])$spearman_cor
knn_spearman_cor(knn_pca_lp[[1]],knn_pca_lp[[2]])$spearman_cor

```

```{r}
# compare the result from manhattan distance under different filtering
knn_spearman_cor(knn_c_s_hvg_lp[[1]],knn_c_s_lp[[1]])$spearman_cor
knn_spearman_cor(knn_c_s_lp[[1]],knn_pca_lp[[1]])$spearman_cor
knn_spearman_cor(knn_c_s_hvg_lp[[1]],knn_pca_lp[[1]])$spearman_cor
```

```{r}
# compare the result from euclidean distance under different filtering
knn_spearman_cor(knn_c_s_hvg_lp[[2]],knn_c_s_lp[[2]])$spearman_cor
knn_spearman_cor(knn_c_s_lp[[2]],knn_pca_lp[[2]])$spearman_cor
knn_spearman_cor(knn_c_s_hvg_lp[[2]],knn_pca_lp[[2]])$spearman_cor
```

## Normalization

```{r}
geodesic_dist_list = list(geod_manhattan_c_s,geod_manhattan_hvg,
                      geod_euclidean_c_s,geod_euclidean_hvg)
names(geodesic_dist_list) = c("Raw Count, Manhattan Distance",
                              "HVG, Manhattan Distance",
                              "Raw Count, Euclidean Distance",
                              "HVG, Euclidean Distance")
```


```{r}
saveRDS(geodesic_dist_list,file = "geodesic_distance_list.RData")
```
```{r}
geodesic_dist_list = readRDS("geodesic_distance_list.RData")

geod_manhattan_c_s = geodesic_dist_list$`Raw Count, Manhattan Distance`
geod_manhattan_hvg = geodesic_dist_list$`HVG, Manhattan Distance`
geod_euclidean_c_s = geodesic_dist_list$`Raw Count, Euclidean Distance`
geod_euclidean_hvg = geodesic_dist_list$`HVG, Euclidean Distance`

```


```{r}
geod_manhattan_hvg[which(geod_manhattan_hvg<quantile(geod_manhattan_hvg,0.025))] = quantile(geod_manhattan_hvg,0.025)
geod_manhattan_c_s[which(geod_manhattan_c_s<quantile(geod_manhattan_c_s,0.025))] = quantile(geod_manhattan_c_s,0.025)
geod_euclidean_hvg[which(geod_euclidean_hvg<quantile(geod_euclidean_hvg,0.025))] = quantile(geod_euclidean_hvg,0.025)
geod_euclidean_c_s[which(geod_euclidean_c_s<quantile(geod_euclidean_c_s,0.025))] = quantile(geod_euclidean_c_s,0.025)
geod_manhattan_hvg[which(geod_manhattan_hvg>quantile(geod_manhattan_hvg,0.975))] = quantile(geod_manhattan_hvg,0.975)
geod_manhattan_c_s[which(geod_manhattan_c_s>quantile(geod_manhattan_c_s,0.975))] = quantile(geod_manhattan_c_s,0.975)
geod_euclidean_hvg[which(geod_euclidean_hvg>quantile(geod_euclidean_hvg,0.975))] = quantile(geod_euclidean_hvg,0.975)
geod_euclidean_c_s[which(geod_euclidean_c_s>quantile(geod_euclidean_c_s,0.975))] = quantile(geod_euclidean_c_s,0.975)
```


```{r}
geod_manhattan_hvg_normed = geod_manhattan_hvg/max(geod_manhattan_hvg)
geod_manhattan_c_s_normed = geod_manhattan_c_s/max(geod_manhattan_c_s)
geod_euclidean_hvg_normed = geod_euclidean_hvg/max(geod_euclidean_hvg)
geod_euclidean_c_s_normed = geod_euclidean_c_s/max(geod_euclidean_c_s)
```

```{r}
range(geod_manhattan_hvg_normed)
range(geod_manhattan_c_s_normed)
range(geod_euclidean_hvg_normed)
range(geod_euclidean_c_s_normed)
```

## Visualize the difference of matrices


```{r}
library(pheatmap)
heatmap1_m = pheatmap::pheatmap(geod_manhattan_c_s_normed, cluster_rows = T, cluster_cols = T,
         name = "geodesic distance\nmanhattan\nwithout filtering",
         color = colorRampPalette(c("white", "red"))(100),
         show_rownames=F,show_colnames=F)
#heatmap1
heatmap2_m = pheatmap::pheatmap(geod_manhattan_hvg_normed, cluster_rows = T, cluster_cols = T,
         name = "geodesic distance\nmanhattan\nafter filtering",
         color = colorRampPalette(c("white", "red"))(100),
         show_rownames=F,show_colnames=F)

geod_manhattan_c_s_cluster <- geod_manhattan_c_s_normed[heatmap1_m$tree_row$order, heatmap1_m$tree_col$order]

geod_manhattan_hvg_cluster <- geod_manhattan_hvg_normed[heatmap1_m$tree_row$order, heatmap1_m$tree_col$order]

pheatmap(geod_manhattan_c_s_cluster,cluster_rows = F, cluster_cols = F,
         name = "geodesic distance\nfrom Manhattan\nwithout filtering",
         color = colorRampPalette(c("white", "red"))(100),
         show_rownames=F,show_colnames=F) 

pheatmap(geod_manhattan_hvg_cluster,cluster_rows = F, cluster_cols = F,
         name = "geodesic distance\nfrom Manhattan\nafter filtering",
         color = colorRampPalette(c("white", "red"))(100),
         show_rownames=F,show_colnames=F)
```

```{r}
library(pheatmap)
heatmap1 = pheatmap::pheatmap(geod_euclidean_c_s_normed, cluster_rows = T, cluster_cols = T,
         name = "geodesic distance\nEuclidean\nwithout filtering",
         color = colorRampPalette(c("white", "red"))(100),
         show_rownames=F,show_colnames=F)
#heatmap1
heatmap2 = pheatmap::pheatmap(geod_euclidean_hvg_normed, cluster_rows = T, cluster_cols = T,
         name = "geodesic distance\nEuclidean\nafter filtering",
         color = colorRampPalette(c("white", "red"))(100),
         show_rownames=F,show_colnames=F)

geod_euclidean_c_s_cluster <- geod_euclidean_c_s_normed[heatmap1$tree_row$order, heatmap1$tree_col$order]

geod_euclidean_hvg_cluster <- geod_euclidean_hvg_normed[heatmap1$tree_row$order, heatmap1$tree_col$order]

pheatmap(geod_euclidean_c_s_cluster,cluster_rows = F, cluster_cols = F,
         name = "geodesic distance\nfrom Euclidean\nwithout filtering",
         color = colorRampPalette(c("white", "red"))(100),
         show_rownames=F,show_colnames=F)

pheatmap(geod_euclidean_hvg_cluster,cluster_rows = F, cluster_cols = F,
         name = "geodesic distance\nfrom Euclidean\nafter filtering",
         color = colorRampPalette(c("white", "red"))(100),
         show_rownames=F,show_colnames=F)
```

### comparing filtering and original

```{r}
diff_m_hvg_vs_cs = geodesic_dist_list$`HVG, Manhattan Distance` - geodesic_dist_list$`Raw Count, Manhattan Distance`
diff_e_hvg_vs_cs = geodesic_dist_list$`HVG, Euclidean Distance` - geodesic_dist_list$`Raw Count, Euclidean Distance`
diff_c_s = geodesic_dist_list$`Raw Count, Manhattan Distance` - geodesic_dist_list$`Raw Count, Euclidean Distance`
diff_hvg = geodesic_dist_list$`HVG, Manhattan Distance` - geodesic_dist_list$`HVG, Euclidean Distance`
```



```{r}
diff_m_hvg_vs_cs = abs(diff_m_hvg_vs_cs)
diff_e_hvg_vs_cs = abs(diff_e_hvg_vs_cs)
diff_m_hvg_vs_cs[which(diff_m_hvg_vs_cs<quantile(diff_m_hvg_vs_cs,0.025))] = quantile(diff_m_hvg_vs_cs,0.025)
diff_e_hvg_vs_cs[which(diff_e_hvg_vs_cs<quantile(diff_e_hvg_vs_cs,0.025))] = quantile(diff_e_hvg_vs_cs,0.025)
diff_m_hvg_vs_cs[which(diff_m_hvg_vs_cs>quantile(diff_m_hvg_vs_cs,0.975))] = quantile(diff_m_hvg_vs_cs,0.975)
diff_e_hvg_vs_cs[which(diff_e_hvg_vs_cs>quantile(diff_e_hvg_vs_cs,0.975))] = quantile(diff_e_hvg_vs_cs,0.975)
```


```{r}
diff_m_hvg_vs_cs_normed = diff_m_hvg_vs_cs/max(diff_m_hvg_vs_cs)
diff_e_hvg_vs_cs_normed = diff_e_hvg_vs_cs/max(diff_e_hvg_vs_cs)
```


```{r}
dif_heatmap1 = pheatmap::pheatmap(diff_m_hvg_vs_cs_normed, cluster_rows = T, cluster_cols = T,
         name = "Difference matrix\n(Manhattan,\nfiltering vs original)",
         color = colorRampPalette(c("white", "red"))(100),
         show_rownames=F,show_colnames=F)
#heatmap1
dif_heatmap2 = pheatmap::pheatmap(diff_e_hvg_vs_cs_normed, cluster_rows = T, cluster_cols = T,
         name = "Difference matrix\n(Euclidean,\nfiltering vs original)",
         color = colorRampPalette(c("white", "red"))(100),
         show_rownames=F,show_colnames=F)

dif_heatmap1_cluster <- diff_m_hvg_vs_cs_normed[dif_heatmap1$tree_row$order, dif_heatmap1$tree_col$order]
dif_heatmap2_cluster <- diff_e_hvg_vs_cs_normed[dif_heatmap1$tree_row$order, dif_heatmap1$tree_col$order]
# normalized difference matrix
pheatmap(dif_heatmap1_cluster, cluster_rows = F, cluster_cols = F,
         name = "Difference matrix\n(Manhattan,\nfiltering vs original)",
         color = colorRampPalette(c("white", "red"))(100),
         show_rownames=F,show_colnames=F)
pheatmap(dif_heatmap2_cluster, cluster_rows = F, cluster_cols = F,
         name = "Difference matrix\n(Euclidean,\nfiltering vs original)",
         color = colorRampPalette(c("white", "red"))(100),
         show_rownames=F,show_colnames=F)
```


### comparing the different distance metrics

```{r}
diff_c_s = abs(diff_c_s)
diff_hvg = abs(diff_hvg)
diff_c_s[which(diff_c_s<quantile(diff_c_s,0.025))] = quantile(diff_c_s,0.025)
diff_hvg[which(diff_hvg<quantile(diff_hvg,0.025))] = quantile(diff_hvg,0.025)
diff_c_s[which(diff_c_s>quantile(diff_c_s,0.975))] = quantile(diff_c_s,0.975)
diff_hvg[which(diff_hvg>quantile(diff_hvg,0.975))] = quantile(diff_hvg,0.975)

diff_c_s_normed = diff_c_s/max(diff_c_s)
diff_hvg_normed = diff_hvg/max(diff_hvg)
```

```{r}
dif_heatmap1 = pheatmap::pheatmap(diff_c_s_normed, cluster_rows = T, cluster_cols = T,
         name = "Difference matrix\n(Manhattan,\nfiltering vs original)",
         color = colorRampPalette(c("white", "red"))(100),
         show_rownames=F,show_colnames=F)
#heatmap1
dif_heatmap2 = pheatmap::pheatmap(diff_hvg_normed, cluster_rows = T, cluster_cols = T,
         name = "Difference matrix\n(Euclidean,\nfiltering vs original)",
         color = colorRampPalette(c("white", "red"))(100),
         show_rownames=F,show_colnames=F)

dif_heatmap1_cluster <- diff_c_s_normed[dif_heatmap1$tree_row$order, dif_heatmap1$tree_col$order]
dif_heatmap2_cluster <- diff_hvg_normed[dif_heatmap1$tree_row$order, dif_heatmap1$tree_col$order]
# normalized difference matrix
pheatmap(dif_heatmap1_cluster, cluster_rows = F, cluster_cols = F,
         name = "Difference matrix\nwithout filtering\n(Manhattan vs Euclidean)",
         color = colorRampPalette(c("white", "red"))(100),
         show_rownames=F,show_colnames=F)
pheatmap(dif_heatmap2_cluster, cluster_rows = F, cluster_cols = F,
         name = "Difference matrix\nafter filtering\n(Manhattan vs Euclidean)",
         color = colorRampPalette(c("white", "red"))(100),
         show_rownames=F,show_colnames=F)
```


### 4 difference matrices use the same order to compare

```{r}
dif_heatmap1 = pheatmap::pheatmap(diff_m_hvg_vs_cs_normed, cluster_rows = T, cluster_cols = T,
         name = "Difference matrix\n(Manhattan,\nfiltering vs original)",
         color = colorRampPalette(c("white", "red"))(100),
         show_rownames=F,show_colnames=F)
#heatmap1
dif_heatmap2 = pheatmap::pheatmap(diff_e_hvg_vs_cs_normed, cluster_rows = T, cluster_cols = T,
         name = "Difference matrix\n(Euclidean,\nfiltering vs original)",
         color = colorRampPalette(c("white", "red"))(100),
         show_rownames=F,show_colnames=F)

dif_heatmap3 = pheatmap::pheatmap(diff_c_s_normed, cluster_rows = T, cluster_cols = T,
         name = "Difference matrix\n(Manhattan,\nfiltering vs original)",
         color = colorRampPalette(c("white", "red"))(100),
         show_rownames=F,show_colnames=F)
#heatmap1
dif_heatmap4 = pheatmap::pheatmap(diff_hvg_normed, cluster_rows = T, cluster_cols = T,
         name = "Difference matrix\n(Euclidean,\nfiltering vs original)",
         color = colorRampPalette(c("white", "red"))(100),
         show_rownames=F,show_colnames=F)

dif_heatmap1_cluster <- diff_m_hvg_vs_cs_normed[dif_heatmap1$tree_row$order, dif_heatmap1$tree_col$order]
dif_heatmap2_cluster <- diff_e_hvg_vs_cs_normed[dif_heatmap1$tree_row$order, dif_heatmap1$tree_col$order]
dif_heatmap3_cluster <- diff_c_s_normed[dif_heatmap1$tree_row$order, dif_heatmap1$tree_col$order]
dif_heatmap4_cluster <- diff_hvg_normed[dif_heatmap1$tree_row$order, dif_heatmap1$tree_col$order]
```


```{r}
# normalized difference matrix
pheatmap(dif_heatmap1_cluster, cluster_rows = F, cluster_cols = F,
         name = "Difference matrix\n(Manhattan,\nfiltering vs original)",
         color = colorRampPalette(c("white", "red"))(100),
         show_rownames=F,show_colnames=F)
pheatmap(dif_heatmap2_cluster, cluster_rows = F, cluster_cols = F,
         name = "Difference matrix\n(Euclidean,\nfiltering vs original)",
         color = colorRampPalette(c("white", "red"))(100),
         show_rownames=F,show_colnames=F)
pheatmap(dif_heatmap3_cluster, cluster_rows = F, cluster_cols = F,
         name = "Difference matrix\nwithout filtering\n(Manhattan vs Euclidean)",
         color = colorRampPalette(c("white", "red"))(100),
         show_rownames=F,show_colnames=F)
pheatmap(dif_heatmap4_cluster, cluster_rows = F, cluster_cols = F,
         name = "Difference matrix\nafter filtering\n(Manhattan vs Euclidean)",
         color = colorRampPalette(c("white", "red"))(100),
         show_rownames=F,show_colnames=F)
```

## correlation plot from sampled entry of matrix
```{r}
set.seed(1)
entry = sample(1:474721, 332305)

geod_manhattan_hvg_normed_sampled = c(geod_manhattan_hvg_normed[entry])
geod_manhattan_c_s_normed_sampled = c(geod_manhattan_c_s_normed[entry])

df_cor_m_hvg_vs_c = data.frame(geod_manhattan_hvg=geod_manhattan_hvg_normed_sampled,
                               geod_manhattan_c_s=geod_manhattan_c_s_normed_sampled)

ggplot(df_cor_m_hvg_vs_c,aes(x=geod_manhattan_hvg, 
                             y=geod_manhattan_c_s))+ 
  geom_point()

```


```{r}
set.seed(1)
entry = sample(1:474721, 332305)

geod_euclidean_hvg_normed_sampled = c(geod_euclidean_hvg_normed[entry])
geod_euclidean_c_s_normed_sampled = c(geod_euclidean_c_s_normed[entry])

# Creating the plot
plot(geod_euclidean_hvg_normed_sampled, geod_euclidean_c_s_normed_sampled, 
     pch = 1, col = "lightblue")

# Regression line
abline(lm(geod_euclidean_hvg_normed_sampled ~ geod_euclidean_c_s_normed_sampled), 
       col = "red", lwd = 3)

df_cor_e_hvg_vs_c = data.frame(geod_euclidean_hvg=geod_euclidean_hvg_normed_sampled,
                               geod_euclidean_c_s=geod_euclidean_c_s_normed_sampled)

ggplot(df_cor_e_hvg_vs_c,aes(x=geod_euclidean_hvg, 
                             y=geod_euclidean_c_s))+ 
  geom_point()


```


```{r fig.width=7,fig.height=5}
df_cor = cbind(df_cor_e_hvg_vs_c,df_cor_m_hvg_vs_c)

pairs(~geod_euclidean_hvg+geod_euclidean_c_s+geod_manhattan_hvg+geod_manhattan_c_s,data=df_cor,
      labels=c("sampled geodesic distance\nfrom euclidean distance metric\nafter filtering",
               "sampled geodesic distance\nfrom euclidean distance metric\nwithout filtering",
               "sampled geodesic distance\nfrom manhattan distance metric\nafter filtering",
               "sampled geodesic distance\nfrom manhattan distance metric\nwithout filtering"))
```

```{r}
cor(df_cor$geod_euclidean_hvg,df_cor$geod_euclidean_c_s,method = "spearman")
cor(df_cor$geod_manhattan_hvg,df_cor$geod_manhattan_c_s,method = "spearman")

cor(df_cor$geod_euclidean_hvg,df_cor$geod_manhattan_hvg,method = "spearman")
cor(df_cor$geod_euclidean_c_s,df_cor$geod_manhattan_c_s,method = "spearman")
```


## Clustering analysis

```{r}
source("DEG_utils.R")
```


```{r}
geodesic_dist_list = readRDS("geodesic_distance_list.RData")

geod_manhattan_c_s = geodesic_dist_list$`Raw Count, Manhattan Distance`
geod_manhattan_hvg = geodesic_dist_list$`HVG, Manhattan Distance`
geod_euclidean_c_s = geodesic_dist_list$`Raw Count, Euclidean Distance`
geod_euclidean_hvg = geodesic_dist_list$`HVG, Euclidean Distance`

geod_manhattan_hvg[which(geod_manhattan_hvg<quantile(geod_manhattan_hvg,0.025))] = quantile(geod_manhattan_hvg,0.025)
geod_manhattan_c_s[which(geod_manhattan_c_s<quantile(geod_manhattan_c_s,0.025))] = quantile(geod_manhattan_c_s,0.025)
geod_euclidean_hvg[which(geod_euclidean_hvg<quantile(geod_euclidean_hvg,0.025))] = quantile(geod_euclidean_hvg,0.025)
geod_euclidean_c_s[which(geod_euclidean_c_s<quantile(geod_euclidean_c_s,0.025))] = quantile(geod_euclidean_c_s,0.025)
geod_manhattan_hvg[which(geod_manhattan_hvg>quantile(geod_manhattan_hvg,0.975))] = quantile(geod_manhattan_hvg,0.975)
geod_manhattan_c_s[which(geod_manhattan_c_s>quantile(geod_manhattan_c_s,0.975))] = quantile(geod_manhattan_c_s,0.975)
geod_euclidean_hvg[which(geod_euclidean_hvg>quantile(geod_euclidean_hvg,0.975))] = quantile(geod_euclidean_hvg,0.975)
geod_euclidean_c_s[which(geod_euclidean_c_s>quantile(geod_euclidean_c_s,0.975))] = quantile(geod_euclidean_c_s,0.975)

geod_manhattan_hvg_normed = geod_manhattan_hvg/max(geod_manhattan_hvg)
geod_manhattan_c_s_normed = geod_manhattan_c_s/max(geod_manhattan_c_s)
geod_euclidean_hvg_normed = geod_euclidean_hvg/max(geod_euclidean_hvg)
geod_euclidean_c_s_normed = geod_euclidean_c_s/max(geod_euclidean_c_s)
```


```{r}
nn_sparseMatrix = function(geodesic_distance_matrix,k=30){
  # first find the names of the 30 nearest neighbor for each cell based on the geodesic distance 
  nn_list = list()
  for (i in 1:nrow(geodesic_distance_matrix)) {
    nn_list[[i]] = names(sort(geodesic_distance_matrix[i,]))[1:k]
  }
  
  # SCT_nn: the matrix contains only 0 and 1
  tmp_nn = geodesic_distance_matrix
  for (i in 1:nrow(geodesic_distance_matrix)) {
    tmp_nn[i,] = 0
    tmp_nn[i,nn_list[[i]]] = 1
  }
  sparsematrix_nn = as(tmp_nn, "sparseMatrix")
  
  # SCT_snn: the nearest neighbor is 1/30, the second nearest is 2/30 etc ...
  # first prepare the 1/30, 2/30, 3/30 ...
  snn = c()
  for (i in 1:k){
    snn[i] = i/k
  }
  # add the names of the nearest neighbor
  snn_list = list()
  for (i in 1:nrow(geodesic_distance_matrix)){
    snn_list[[i]] = snn
    names(snn_list[[i]]) = nn_list[[i]]
  }
  # put the values (1/30, 2/30, 3/30 ...) into the matrix
  tmp_snn = geodesic_distance_matrix
  for (i in 1:nrow(geodesic_distance_matrix)) {
    tmp_snn[i,] = 0
    tmp_snn[i,nn_list[[i]]] = snn_list[[i]]
  }
  sparsematrix_snn = as(tmp_snn, "sparseMatrix")

  return(list(SCT_nn = sparsematrix_nn,
              SCT_snn = sparsematrix_snn))

}
```

```{r}
nn_euclidean_s = nn_sparseMatrix(geod_euclidean_c_s_normed)[[1]]
snn_euclidean_s = nn_sparseMatrix(geod_euclidean_c_s_normed)[[2]]
nn_euclidean_hvg = nn_sparseMatrix(geod_euclidean_hvg_normed)[[1]]
snn_euclidean_hvg = nn_sparseMatrix(geod_euclidean_hvg_normed)[[2]]
nn_manhattan_s = nn_sparseMatrix(geod_manhattan_c_s_normed)[[1]]
snn_manhattan_s = nn_sparseMatrix(geod_manhattan_c_s_normed)[[2]]
nn_manhattan_hvg = nn_sparseMatrix(geod_manhattan_hvg_normed)[[1]]
snn_manhattan_hvg = nn_sparseMatrix(geod_manhattan_hvg_normed)[[2]]
```


```{r}
find_seurat_clusters_nn <-
    function(seurat_obj,# with our nearest neighbor
             findPC_npcs = 100,
             pcs = NULL,
             clustering_resolution = 0.5,
             redo_knn = FALSE,
             verbose = FALSE) {
        
      seurat_obj <- FindClusters(seurat_obj,#graph.name = "SCT_nn",
                                   resolution = clustering_resolution,
                                   verbose = verbose)
        
        seurat_clusters = as.numeric(as.character(seurat_obj$seurat_clusters)) + 1
        Idents(seurat_obj) <- as.factor(seurat_clusters)
        
        list(
            seurat_clusters = seurat_clusters,
            n_clusters = max(seurat_clusters),
            clustering_resolution = clustering_resolution,
            pcs = pcs,
            seurat_obj = seurat_obj,
            knn_graph = seurat_obj@graphs$RNA_snn
        )
    }
```



```{r}
c_s = readRDS("c_s.RData")
seurat_object_s = CreateSeuratObject(c_s)
```

```{r}
tmp = find_seurat_clusters(seurat_object_s)

tmp$seurat_obj
```

```{r}
seurat_object_s_e = seurat_object_s
seurat_object_s_e@graphs$RNA_nn = as.Graph(nn_euclidean_s)
seurat_object_s_e@graphs$RNA_snn = as.Graph(snn_euclidean_s)

clusters_result_s_e = find_seurat_clusters_nn(seurat_object_s_e)
clusters_result_s_e$seurat_obj
clusters_result_s_e$n_clusters

seurat_object_s_m = seurat_object_s
seurat_object_s_m@graphs$RNA_nn = as.Graph(nn_manhattan_s)
seurat_object_s_m@graphs$RNA_snn = as.Graph(snn_manhattan_s)

clusters_result_s_m = find_seurat_clusters_nn(seurat_object_s_m)
clusters_result_s_m$seurat_obj
clusters_result_s_m$n_clusters

```


```{r}
seurat_object_hvg_e = seurat_object
seurat_object_hvg_e@graphs$RNA_nn = as.Graph(nn_euclidean_hvg)
seurat_object_hvg_e@graphs$RNA_snn = as.Graph(snn_euclidean_hvg)

clusters_result_hvg_e = find_seurat_clusters_nn(seurat_object_hvg_e)
clusters_result_hvg_e$seurat_obj
clusters_result_hvg_e$n_clusters

seurat_object_hvg_m = seurat_object
seurat_object_hvg_m@graphs$RNA_nn = as.Graph(nn_manhattan_hvg)
seurat_object_hvg_m@graphs$RNA_snn = as.Graph(snn_manhattan_hvg)

clusters_result_hvg_m = find_seurat_clusters_nn(seurat_object_hvg_m)
clusters_result_hvg_m$seurat_obj
clusters_result_hvg_m$n_clusters
```

```{r}
resolution = 0.5
pcs = 20
nfeatures = 2000
verbose = FALSE

seurat_object = CreateSeuratObject(c_s)

seurat_object <- NormalizeData(seurat_object)
seurat_object <- FindVariableFeatures(seurat_object,
                                        selection.method = "vst",
                                        nfeatures = nfeatures)

seurat_object <- ScaleData(seurat_object,
                            features = rownames(seurat_object))

seurat_object <- RunPCA(seurat_object,
                        features = VariableFeatures(object = seurat_object), 
                        npcs = 50,
                        verbose = verbose)

# run clustering and dim reduction
seurat_object <- FindNeighbors(seurat_object, dims = 1:pcs,verbose = verbose)
clusters_result_standard = find_seurat_clusters_nn(seurat_object)
clusters_result_standard$n_clusters
clusters_result_standard$seurat_obj
```

```{r}
resolution = 0.5
pcs = 20
nfeatures = 36601
verbose = FALSE

seurat_object_unfilter = CreateSeuratObject(c_s)

seurat_object_unfilter <- NormalizeData(seurat_object_unfilter)
seurat_object_unfilter <- FindVariableFeatures(seurat_object_unfilter,
                                        selection.method = "vst",
                                        nfeatures = nfeatures)

seurat_object_unfilter <- ScaleData(seurat_object_unfilter,
                            features = rownames(seurat_object_unfilter))

seurat_object_unfilter <- RunPCA(seurat_object_unfilter,
                        features = VariableFeatures(object = seurat_object_unfilter), 
                        npcs = 50,
                        verbose = verbose)

# run clustering and dim reduction
seurat_object_unfilter <- FindNeighbors(seurat_object_unfilter, dims = 1:pcs,verbose = verbose)
clusters_result_standard_unfilter = find_seurat_clusters_nn(seurat_object_unfilter)
clusters_result_standard_unfilter$seurat_obj
clusters_result_standard_unfilter$n_clusters
```




```{r}
seurat_clusters_nn <-
    function(seurat_obj,# with our nearest neighbor
             n_clusters = NULL,
             pcs = NULL,
             start_resolution = 0.5,
             step_size = 0.1,
             redo_knn = FALSE,
             verbose = FALSE) {
        clusters_result <- find_seurat_clusters_nn(
            seurat_obj,
            pcs = pcs,
            clustering_resolution = start_resolution,
            redo_knn = redo_knn,
            verbose = verbose
        )
        
        # if we don't need to tune clustering_resolution, just return
        if (is.null(n_clusters)) {
            return(clusters_result)
        }
        
        # if we need to have a fixed # of clusters, tune the clustering_resolution to achieve this
        tuned_resolution <- start_resolution
        last_nclusters <- clusters_result$n_clusters
        while (last_nclusters != n_clusters) {
            if (last_nclusters > n_clusters) {
                tuned_resolution <- round(tuned_resolution - step_size, 3)
            } else {
                tuned_resolution <- round(tuned_resolution + step_size, 3)
            }
            clusters_result <-
                find_seurat_clusters_nn(
                    seurat_obj = clusters_result$seurat_obj,
                    pcs = pcs,
                    clustering_resolution = tuned_resolution,
                    verbose = verbose
                )
            
            if ((last_nclusters > n_clusters) != (clusters_result$n_clusters > n_clusters)) {
                step_size <- round(step_size / 1.5, 3)
            }
            
            # avoid infinit loop
            last_nclusters <-
                length(unique(clusters_result$seurat_clusters))
        }
        clusters_result
    }
```

```{r}
standard_cluster_result = seurat_clusters_nn(seurat_object,n_clusters = 6)
hvf_e_cluster_result = seurat_clusters_nn(seurat_object_hvg_e,n_clusters = 6)
hvf_m_cluster_result = seurat_clusters_nn(seurat_object_hvg_m,n_clusters = 6)
standard_unfilter_cluster_result = seurat_clusters_nn(seurat_object_unfilter,n_clusters = 6)
unfilter_m_cluster_result = seurat_clusters_nn(seurat_object_s_m,n_clusters = 6)
unfilter_e_cluster_result = seurat_clusters_nn(seurat_object_s_e,n_clusters = 6)
```


```{r}
standard_cluster_result
hvf_e_cluster_result
hvf_m_cluster_result
standard_unfilter_cluster_result
unfilter_m_cluster_result
unfilter_e_cluster_result
```


```{r}
adjustedRandIndex(standard_cluster_result$seurat_clusters,
                  standard_unfilter_cluster_result$seurat_clusters)
```
```{r}
adjustedRandIndex(standard_cluster_result$seurat_clusters,
                  hvf_e_cluster_result$seurat_clusters)
adjustedRandIndex(standard_cluster_result$seurat_clusters,
                  hvf_m_cluster_result$seurat_clusters)
adjustedRandIndex(standard_cluster_result$seurat_clusters,
                  unfilter_e_cluster_result$seurat_clusters)
adjustedRandIndex(standard_cluster_result$seurat_clusters,
                  unfilter_m_cluster_result$seurat_clusters)
```


```{r}
adjustedRandIndex(standard_unfilter_cluster_result$seurat_clusters,
                  hvf_e_cluster_result$seurat_clusters)
adjustedRandIndex(standard_unfilter_cluster_result$seurat_clusters,
                  hvf_m_cluster_result$seurat_clusters)
adjustedRandIndex(standard_unfilter_cluster_result$seurat_clusters,
                  unfilter_e_cluster_result$seurat_clusters)
adjustedRandIndex(standard_unfilter_cluster_result$seurat_clusters,
                  unfilter_m_cluster_result$seurat_clusters)
```







